---
title: "Economic and Health Effects of Environmental Events"
output: html_document
---

#Economic and Health Effects of Environmental Events

##Summary

##Data Processing

First, we unzip the file and load the .csv.library

```{r}
StormData <- read.csv("./repdata-data-StormData.csv.bz2")
```


Next, we will calculate the primary metric by which we will determine damage to population health (fatalities + injuries)



```{r}
library(dplyr)
StormData <- mutate(StormData,TOTAL_POP_DMG = FATALITIES + INJURIES)
```

We will also calculate the primary metric by which we will calculate economic damage (property damage + crop damage). In order to do this, we need to first convert the exponent columns to numeric values.

There are several different values in both the PROPDMGEXP and CROPDMGEXP columns.
We will be using this method[https://rstudio-pubs-static.s3.amazonaws.com/58957_37b6723ee52b455990e149edde45e5b6.html] for creating the correct exponent values.

```{r}
StormData$PROPDMGEXP <- gsub("0",1,StormData$PROPDMGEXP,ignore.case = TRUE)
StormData$PROPDMGEXP <- gsub("1",1,StormData$PROPDMGEXP,ignore.case = TRUE)
StormData$PROPDMGEXP <- gsub("2",1,StormData$PROPDMGEXP,ignore.case = TRUE)
StormData$PROPDMGEXP <- gsub("3",1,StormData$PROPDMGEXP,ignore.case = TRUE)
StormData$PROPDMGEXP <- gsub("4",1,StormData$PROPDMGEXP,ignore.case = TRUE)
StormData$PROPDMGEXP <- gsub("5",1,StormData$PROPDMGEXP,ignore.case = TRUE)
StormData$PROPDMGEXP <- gsub("6",1,StormData$PROPDMGEXP,ignore.case = TRUE)
StormData$PROPDMGEXP <- gsub("7",1,StormData$PROPDMGEXP,ignore.case = TRUE)
StormData$PROPDMGEXP <- gsub("8",1,StormData$PROPDMGEXP,ignore.case = TRUE)
StormData$PROPDMGEXP <- gsub("\\+",1,StormData$PROPDMGEXP,ignore.case = TRUE)

StormData$PROPDMGEXP <- gsub("\\-",0,StormData$PROPDMGEXP,ignore.case = TRUE)
StormData$PROPDMGEXP <- gsub("\\?",0,StormData$PROPDMGEXP,ignore.case = TRUE)

StormData$PROPDMGEXP <- gsub("H",2,StormData$PROPDMGEXP,ignore.case = TRUE)
StormData$PROPDMGEXP <- gsub("K",3,StormData$PROPDMGEXP,ignore.case = TRUE)
StormData$PROPDMGEXP <- gsub("M",6,StormData$PROPDMGEXP,ignore.case = TRUE)
StormData$PROPDMGEXP <- gsub("B",9,StormData$PROPDMGEXP,ignore.case = TRUE)

StormData$CROPDMGEXP <- gsub("0",1,StormData$CROPDMGEXP,ignore.case = TRUE)
StormData$CROPDMGEXP <- gsub("2",1,StormData$CROPDMGEXP,ignore.case = TRUE)

StormData$CROPDMGEXP <- gsub("\\?",0,StormData$CROPDMGEXP,ignore.case = TRUE)

StormData$CROPDMGEXP <- gsub("H",2,StormData$CROPDMGEXP,ignore.case = TRUE)
StormData$CROPDMGEXP <- gsub("K",3,StormData$CROPDMGEXP,ignore.case = TRUE)
StormData$CROPDMGEXP <- gsub("M",6,StormData$CROPDMGEXP,ignore.case = TRUE)
StormData$CROPDMGEXP <- gsub("B",9,StormData$CROPDMGEXP,ignore.case = TRUE)

StormData$CROPDMGEXP <- as.numeric(StormData$CROPDMGEXP)
StormData$PROPDMGEXP <- as.numeric(StormData$PROPDMGEXP)

StormData$PROPDMGEXP<-replace(StormData$PROPDMGEXP,is.na(StormData$PROPDMGEXP),0)
StormData$CROPDMGEXP<-replace(StormData$CROPDMGEXP,is.na(StormData$CROPDMGEXP),0)
```

Then, we will calculate the total economic damage ((PROPDMG * 10^PROPDMGEXP) + (CROPDMG * 10^CROPDMGEXP)).

```{r}
StormData <- mutate(StormData,TOTAL_ECON_DMG = (PROPDMG * 10^PROPDMGEXP) + (CROPDMG * 10^CROPDMGEXP))
```

Finally, we'll group by EVTYPE

```{r}
StormData <- group_by(StormData,EVTYPE)
```

##Results

Tornadoes have the highest human cost by far. Total fatalities + injuries surpass the next highest event (Excessive Heat) by more than 10x. In addition, two other wind-related events appear in the top 10.
```{r}
POP_DMG <- summarise(StormData,TOTAL_FATALITIES = sum(FATALITIES),TOTAL_INJURIES = sum(INJURIES),TOTAL_POP_DMG = sum(TOTAL_POP_DMG))
arrange(POP_DMG,desc(TOTAL_POP_DMG))
```

While Tornadoes make an appearance in the top 10 events in terms of economic impact, Floods cause almost 3x as much damage. In fact, 6 of the top 10 most economically devastating events are floods or hurricanes (where are a great deal of the damage is caused by flooding).
```{r}
ECON_DMG <- summarise(StormData,TOTAL_PROP = sum(PROPDMG * 10^PROPDMGEXP), TOTAL_CROP = sum(CROPDMG * 10^CROPDMGEXP), TOTAL_ECON_DMG = sum(TOTAL_ECON_DMG) )
arrange(ECON_DMG,desc(TOTAL_ECON_DMG))
```


Finally, let's look at Total Economic and Population damage over time
```{r}
DMG_BY_YEAR <- mutate(StormData,YEAR = year(strptime(BGN_DATE,format = "%m/%d/%Y %H:%M:%S"))) %>% group_by(YEAR) %>% summarise(TOTAL_ECON_DMG = sum(TOTAL_ECON_DMG),TOTAL_POP_DMG = sum(TOTAL_POP_DMG))


```
